Describing demographic models is hard

### 2020-05-20 paper.tex
Pass through the bad model section.

To illustrate the demography API, \msprime\ included a description of a widely-used
three population Out-of-Africa model~\citep{gutenkunst2009inferring}
as part of its tutorial documentation. In this model (Fig.~\ref{fig:ooa_stats}A),
Eurasian (CEU and CHB) and African (YRI) populations split from each other in the deep past,
followed by a more recent split of European and Asian populations, with variable rates of
continuous migration between each of the populations. However, the implementation in the
\msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
populations, when there should have been just a single randomly mating population, migration was
allowed to occur between the ancestral population and a second population with size equal to
the Eurasian bottleneck size for all time into the past (Fig.~\ref{fig:ooa_stats}B).

This model was subsequently copied
several times and used in publications. By searching for some keywords on
GitHub, we found 33 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repos to alert the authors.)

It was used in a number of publications
~\citep{martin2017human,albers2020dating,kelleher2019inferring,tong2020population,
overcast2019integrated,zhou2018popdemog,pfaffelhuber2020choose}

Fortunately, the effects of this error are subtle. Population sizes and structure since the time of
the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
the correct and incorrect model. However, the ancient structure distorts the distribution
of $T_{MRCA}$ beyond the split date and increases
heterozygosity in contemporary populations roughly $4\%$ over expectations
from the correct model (Fig.~\ref{fig:ooa_stats}E-F).

%However, the presence of the ghost population increases
%the effective population size \aprcomment{not true - the effect is a bit different than just change in Ne}
%and thus affects the distribution of older $T_{MRCA}$ and the expected
%heterozygosity in each of the three populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).

Such errors can be prevented by improved API design choice.
To model a population split currently in \msprime, a user must specify a
\texttt{MassMigration} event that moves lineages from one population to another
and then must also remember to turn off migration
between those populations at the same time.
The release of \msprime~ver.~1.0 will introduce a \texttt{PopulationSplit} event,
which more intuitively links the merger of lineages with appropriate changes in
migration rates at the time of the split.

@@

We computed the expected allele frequency spectrum using
\texttt{moments}~ver.~1.0.3~\citep{jouganous2017inferring}, and LD-decay curves using
\texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other diversity statistics
were computed from the expected AFS and verified using branch statistics from the
tree sequence records of \msprime\ simulations~\citep{ralph2020efficiently}.
Demographic models were plotted using the \texttt{demography} package
(\url{https://github.com/apragsdale/demography}, ver.~0.0.3).
We used the original pipeline from~\citet{martin2017human} available from
\url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py}.
We updated the pipeline to run with the correct demographic parameters and more
recent versions of \msprime\ and \tskit\, and the updated pipeline is
available at \url{https://github.com/apragsdale/PRS}.
Data and python scripts to recreate Figures~1~and~2 can be found at
\url{https://github.com/jeromekelleher/msprime-model-errors}.

>>>>

To illustrate the demography API, \msprime\ included a description of a widely-used
three population Out-of-Africa model~\citep{gutenkunst2009inferring}
as part of its tutorial documentation. In this model (Fig.~\ref{fig:ooa_stats}A),
Eurasian (CEU and CHB) and African (YRI) populations split from each other in the deep past,
followed by a more recent split of European and Asian populations, with variable rates of
continuous migration between each of the populations. However, the implementation in the
\msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
populations, when there should have been just a single randomly mating population, migration was
allowed to occur between the ancestral population and a second population with size equal to
the Eurasian bottleneck size for all time into the past
(Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
the tutorial for \msprime\ version 0.3.0 and remained in the documentation for
around four years.

Fortunately, the effects of this error are subtle and it would take
and exceedingly sensitive method to detect the difference.
Population sizes and structure since the time of
the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
the correct and incorrect model. However, the ancient structure distorts the distribution
of $T_{MRCA}$ beyond the split date and increases
heterozygosity in contemporary populations roughly $4\%$ over expectations
from the correct model (Fig.~\ref{fig:ooa_stats}E-F).

It is fortunate that the error in the model description has quite a
small effect because the tutorial code has been copied many times
and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repos to alert the authors.)
In most cases the publications use simulations to test some inference
method which is not directly concerned with detailed demography,
and the model is simply used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} use the incorrect model as an example
of their method for visualising demographic models can support
\msprime\ input.
% TODO ask Franz about this
Finally, \cite{pfaffelhuber2020choose}
use simulations of the incorrect model demography to evalute
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
will have any qualitative effect on their conclusions.

This long-standing error could have been prevented by better API design.
To model a population split currently in \msprime, a user must specify a
\texttt{MassMigration} event that moves lineages from one population to another
and then must also remember to turn off migration
between those populations at the same time.
The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
which more intuitively links the merger of lineages with appropriate changes in
migration rates at the time of the split.

@@

We computed the expected allele frequency spectrum using
\texttt{moments}~ver.~1.0.3~\citep{jouganous2017inferring}, and LD-decay curves using
\texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other diversity statistics
were computed from the expected AFS and verified using branch statistics from the
tree sequence records of \msprime\ simulations~\citep{ralph2020efficiently}.
Demographic models were plotted using the \texttt{demography} package
(\url{https://github.com/apragsdale/demography}, ver.~0.0.3).
We used the original pipeline from~\citet{martin2017human} available from
\url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py}.
We updated the pipeline to run with the correct demographic parameters and more
recent versions of \msprime\ and \tskit\, and the updated pipeline is
available at \url{https://github.com/apragsdale/PRS}.
Data and python scripts to recreate Figures~1~and~2 can be found at
\url{https://github.com/jeromekelleher/msprime-model-errors}.
A full list of the GitHub repositories containing copies of the erroneous
model are also given here.
